---
title: "Thesis work"
author: "Loubna Guenoune"
date: "2024-05-28"
output: html_document
---
#Chapter two:
cholera map
```{r}
library(cholera)
rng <- mapRange()
plot(fatalities[, c("x", "y")],
  pch = 15, col = "black",
  cex = 0.4, xlim = rng$x, ylim = rng$y, asp = 1,
  frame.plot = TRUE, axes = FALSE
)
addRoads()
```

spatial neighbourhood distance measure map :

```{r}
nb <- dnearneigh(x = st_centroid(map), d1 = 0, d2 = 0.4)
plot(st_geometry(map), border = "lightgray")
plot.nb(nb, st_geometry(map), add = TRUE)
```

#Chapter 3 :


```{r}
library(tidyverse)
library(sf)
library(INLA)
library(RColorBrewer)
library(tmap)
library(patchwork)
library(rnaturalearth)
library(terra)
library(ggspatial)
library(rnaturalearthdata)
```

```{r}
library(readxl)
final_data <-read_excel("C:/Users/Lenovo/Desktop/rédaction/Dataiowa.xlsx")
```

```{r}
final_data$Corn_Yield<-as.numeric(final_data$Corn_Yield)

```
#Plotting the Corn yield for the whole state over the years :
```{r}
Data1 <- final_data %>%
  group_by(YEAR) %>%
  summarize(Corn_Yield = mean(as.numeric(Corn_Yield), na.rm = TRUE))  # Maybe use mean instead of sum

ggplot(data = Data1, aes(x = YEAR, y = Corn_Yield)) +
  # Theme with lighter background and grid
  theme_bw() +
  theme(
    panel.background = element_rect(fill = "white"),
    panel.grid.major.x = element_line(color = "gray80", linetype = "dashed"),
    panel.grid.major.y = element_line(color = "gray80", linetype = "dashed")
  ) +
  # Customize line color and size
  geom_line(color = "#377EB8", size = 1.2) +  # Adjust color (blue) and line thickness
  # Add labels and title
  labs(title = "Average Corn Yield in Iowa Over Time",
       x = "Year",
       y = "Corn Yield ") +
  # Adjust x and y-axis appearance
  scale_x_continuous(breaks = seq(min(Data1$YEAR), max(Data1$YEAR), by = 1)) +  # Adjust x-axis breaks for readability
  scale_y_continuous() +
  # Add theme elements (optional)
  theme(plot.title = element_text(hjust = 0.5))  # Center title

```

```{r}

library(GGally)
# Customize the pairwise plot
p <- ggpairs(
  data = final_data, 
  columns = c("Corn_Yield","Temp","precipitation","Hum","dlh","wind","DNI"),
  upper = list(continuous = wrap("cor", size = 4, color = "darkslategray")),
  lower = list(continuous = wrap("smooth", color = "steelblue")),
  diag = list(continuous = wrap("barDiag", fill = "darkslategray"))
) +
  theme_bw() +
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    panel.grid.major = element_line(color = "gray90"),
    panel.grid.minor = element_blank(),
    axis.text = element_text(color = "darkslategray"),
    axis.title = element_text(color = "darkslategray"),
    strip.background = element_rect(fill = "white", color = "gray70"),
    strip.text = element_text(color = "darkslategray")
  )

# Save the plot with appropriate size for a Word document
ggsave("pairwise_plot_classy.png", plot = p, width = 10, height = 7, units = "in", dpi = 300)
p
```



```{r}
library(tigris)
library(sf)
options(tigris_use_cache = TRUE)

temp_dir <- tempdir()
options(tempdir = temp_dir)

iowa_counties <- counties(state = "IA", class = "sf")
print(iowa_counties)
iowa_crs <- st_crs(iowa_counties)

```


```{r}
final_data$County[final_data$County == "Carrol"] <- "Carroll"
final_data$County[final_data$County == "Monrtgomery"] <- "Montgomery"
iowa_counties$NAME[iowa_counties$NAME=="O'Brien"]<-"Obrien"
```

```{r}
ggplot() + 
  geom_sf(data = iowa_counties,col = "navy" , fill = "white") + 
  theme_bw() +
  geom_sf_text(data = iowa_counties,mapping = aes(label = NAME),size = 1.2) +
  ggtitle("Iowa")

```
#Spatial Neighbours and spatial Neighbourhood Graph :
```{r}
library(sp)
library(geosphere)
library(spdep)
# Convert sf object to Spatial object for compatibility with spdep functions
full_poly <- as_Spatial(iowa_counties)

# Calculate the centroids of the counties (using sf functions)
centroids <- st_centroid(iowa_counties)

# Extract coordinates of the centroids
centroid_coords <- st_coordinates(centroids)

# Compute the pairwise distance matrix 
dist_matrix <- distm(centroid_coords, fun = distHaversine)

# Set the distance threshold 
distance_threshold <- 60  # Adjust to the required distance



# Create a neighbor list based on the distance threshold
neighbors <- dnearneigh(centroid_coords, 0, distance_threshold, longlat = TRUE)

# Export neighbor list to INLA format
nb2INLA("iowa.graph", neighbors)

# Generate file path for the adjacency graph
iowa.adj <- paste(getwd(), "/iowa.graph", sep = "")

# Convert neighbor list to spatial weights list
wmat <- nb2listw(neighbors, style = "B", zero.policy = TRUE)

# Plot the polygon map
plot(st_geometry(iowa_counties), border = "navy")

# Add neighbor links to the plot
plot(wmat, centroid_coords, pch = 19, cex = 0.6, add = TRUE, col = "maroon")

```

Merging Data :

```{r}
iowa_counties <- iowa_counties %>%
  left_join(final_data, by = c( "NAME"="County" ))
iowa_counties
```

```{r}
iowa_counties <- iowa_counties %>%
  rename( County=NAME)
```

checkup on neighbors: 

```{r}
summary(neighbors)

```

```{r}
# Load necessary libraries
library(sf)
library(tmap)
library(dplyr)
library(RColorBrewer)


# Step 1: Aggregate corn yield data by county over the 14 years
aggregated_corn_yield <- iowa_counties%>%
  group_by(NAME) %>%
  summarize(Total_Corn_Yield = sum(Corn_Yield, na.rm = TRUE))


# Step 3: Set the tmap mode to plotting
tmap_mode("plot")

# Step 4: Create the map with a darker color palette
corn_yield_map <- tm_shape(aggregated_corn_yield) +
  tm_polygons("Total_Corn_Yield", 
              title = "Total Corn Yield ",
              palette = "YlOrBr", 
              style = "quantile",
              border.col = "gray40",  # Dark gray borders for counties
              border.alpha = 0.5) +  # Semi-transparent borders
  tm_layout(
    title = "Total Corn Yield in Iowa by County (2010-2023)",
    title.position = c("center", "top"),
    legend.outside = TRUE,
    legend.title.size = 1.2,
    legend.text.size = 0.8,
    legend.bg.color = "white",
    legend.bg.alpha = 0.8,
    frame = FALSE,  # No frame around the map
    bg.color = "white",  # White background
    inner.margins = c(0.1, 0.1, 0.1, 0.1),  # Margins for title and legend
    fontfamily = "sans"  # Font family for text
  )

# Step 5: Print the map
print(corn_yield_map)

# Step 6: Save the map as an image file
tmap_save(corn_yield_map, "iowa_corn_yield_map_classy.png", width = 10, height = 7, units = "in", dpi = 300)

```




Moran's I and Geary's C, for the sum of corn yield over the years :
Moran's I:
```{r}
aggregated_data <- iowa_counties %>%
  group_by(County) %>%
  summarize(Total_Corn_Yield = sum(Corn_Yield, na.rm = TRUE))

morans_i <- moran.test(aggregated_data$Total_Corn_Yield, wmat)
morans_i

moran.plot(x = aggregated_data$Total_Corn_Yield,listw = wmat,quiet = F)

```


```{r}
# Load necessary libraries
library(spdep)
library(dplyr)
library(ggplot2)
library(ggrepel)


# Compute the spatially lagged values
lagged_values <- lag.listw(wmat, aggregated_data$Total_Corn_Yield)

# Create a data frame for plotting
plot_data <- data.frame(
  Total_Corn_Yield = aggregated_data$Total_Corn_Yield,
  Lagged_Values = lagged_values,
  County = aggregated_data$County
)



# Create the Moran scatterplot using ggplot2
moran_plot<-ggplot(plot_data, aes(x = Total_Corn_Yield, y = Lagged_Values, label = County)) +
  geom_point(color = "navy", size = 3) +
  geom_text_repel(size = 3) + # Using ggrepel for better text positioning
  geom_smooth(method = "lm", color = "maroon", se = FALSE) + # Add the regression line
  geom_vline(xintercept = mean(plot_data$Total_Corn_Yield), linetype = "dashed", color = "grey") + # Mean line
  geom_hline(yintercept = mean(plot_data$Lagged_Values), linetype = "dashed", color = "grey") + # Mean line
  labs(
    title = "Moran Scatterplot",
    x = "Total Corn Yield",
    y = "Spatial Lag of Total Corn Yield"
  ) +
  theme_minimal(base_size = 10) + # Use a minimal theme for a clean look
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"), # Center and bold the title
    axis.title = element_text(face = "bold"), # Bold the axis titles
    panel.grid.major = element_line(color = "grey", size = 0.2), # Light grid lines
    panel.grid.minor = element_blank() # No minor grid lines
  )

```

Geary's C :

```{r}
gearys_c <- geary.test(aggregated_data$Total_Corn_Yield, wmat)
gearys_c
```
##Local Moran's I :


```{r}
library(dplyr)
library(spdep)
library(sf)
library(ggplot2)


# Step 1: Aggregate corn yield by county across all years
aggregated_data <- final_data %>%
  group_by(County) %>%
  summarize(Total_Corn_Yield = sum(Corn_Yield, na.rm = TRUE))

# Ensure that the order of rows in aggregated_data matches the spatial weights matrix nb
aggregated_data <- aggregated_data %>%
  arrange(County)

# Step 2: Calculate Local Moran's I
lw1 <- nb2listw(neighbors, style="W")
local_morans_i <- localmoran(aggregated_data$Total_Corn_Yield, lw1)

# Convert local_morans_i to a data frame
local_morans_i_df <- as.data.frame(local_morans_i)

# Add a column with the County names
local_morans_i_df$County <- aggregated_data$County

# Step 3: Merge Local Moran's I results with the sf object
iowa_counties2 <- iowa_counties %>%
  left_join(local_morans_i_df, by = c("County"="County"))



# Define a custom color palette
breaks <- c(-Inf, -0.5, -0.25, 0, 0.25, 0.5, Inf)

# Define labels for each bin
labels <- c("Very Negative", "Negative", "Slightly Negative", "Slightly Positive", "Positive", "Very Positive")

# Create a new variable representing the bins
iowa_counties2$Ii_bin <- cut(iowa_counties2$Ii, breaks = breaks, labels = labels)

# Define a custom color palette for each category
custom_palette <- c("#2B3D4F", "#4F628E", "#6C7691", "#A4838D", "#882D43", "#C66B6E")

map_plot<-ggplot(data = iowa_counties2) +
  geom_sf(aes(fill = Ii_bin), color = "gray30", size = 0.1) +  # Add a thin gray border around counties
  scale_fill_manual(values = setNames(custom_palette, labels), name = "Local Moran's I") +  # Use custom color palette
  theme_minimal() +
  theme(legend.position = "left",  # Move legend to the bottom
        legend.key.height = unit(1, "cm"),  # Increase legend key height
        legend.text = element_text(size = 10),  # Adjust legend text size
        plot.title = element_text(size = 14, face = "bold", hjust = 0.5)) +  # Adjust title
  labs(title = "Local Moran's I for Corn Yield in Iowa")

ggsave("locmoran.png", plot = map_plot,width=6,height=5 ,dpi = 300)
```




```{r}
# Define breaks for binning the p-values
breaks <- c(-Inf, 0.05, Inf)

# Define labels for each bin
labels <- c("Less than 0.05", "Above 0.05")

# Create a new variable representing the bins
iowa_counties2$`Pr(z != E(Ii))`<-as.numeric(iowa_counties2$`Pr(z != E(Ii))`)
iowa_counties2$P_Value_Bin <- cut(iowa_counties2$`Pr(z != E(Ii))`, breaks = breaks, labels = labels)

# Define a custom color palette for each category
custom_palette <- c("#2B3D4F", "#C66B6E")  # Adjust colors if needed

map_plot <- ggplot(data = iowa_counties2) +
  geom_sf(aes(fill = P_Value_Bin), color = "gray30", size = 0.1) +  # Add a thin gray border around counties
  scale_fill_manual(values = setNames(custom_palette, labels), name = "P-Value of LMI") +  # Use custom color palette
  theme_minimal() +
  theme(legend.position = "left",  # Move legend to the bottom
        legend.key.height = unit(0.7, "cm"),  # Increase legend key height
        legend.text = element_text(size = 6),  # Adjust legend text size
        plot.title = element_text(size = 8, face = "bold", hjust = 0.5)) +  # Adjust title
  labs(title = "P-Value of Local Moran's I for Corn Yield in Iowa")  # Adjust title if needed

ggsave("locmoran.png", plot = map_plot, width = 6, height = 5, dpi = 300)
map_plot
```

```{r}
library(gridExtra)

# Define breaks for binning the p-values
breaks_p <- c(-Inf, 0.05, Inf)

# Define labels for each bin
labels_p <- c("Less than 0.05", "Above 0.05")

# Create a new variable representing the bins for p-values
iowa_counties2$P_Value_Bin <- cut(iowa_counties2$`Pr(z != E(Ii))`, breaks = breaks_p, labels = labels_p)

# Define a custom color palette for each category of p-values
custom_palette_p <- c("#2B3D4F", "#C66B6E")  # Adjust colors if needed

# Map for p-values
map_plot_p <- ggplot(data = iowa_counties2) +
  geom_sf(aes(fill = P_Value_Bin), color = "gray30", size = 0.1) +  # Add a thin gray border around counties
  scale_fill_manual(values = setNames(custom_palette_p, labels_p), name = "P-Value LMI") +  # Use custom color palette
  theme_minimal() +
  theme(legend.position = "left",  # Move legend to the bottom
        legend.key.height = unit(0.5, "cm"),  # Increase legend key height
        legend.text = element_text(size = 6),  # Adjust legend text size
        plot.title = element_text(size = 8, face = "bold", hjust = 0.5)) +  # Adjust title
  labs(title = "P-Value of Local Moran's I for Corn Yield in Iowa")  # Adjust title if needed

# Define breaks for binning the Moran's I values
breaks_m <- c(-Inf, -0.5, -0.25, 0, 0.25, 0.5, Inf)

# Define labels for each bin
labels_m <- c("Very Negative", "Negative", "Slightly Negative", "Slightly Positive", "Positive", "Very Positive")

# Create a new variable representing the bins for Moran's I values
iowa_counties2$Ii_Bin <- cut(iowa_counties2$Ii, breaks = breaks_m, labels = labels_m)

# Define a custom color palette for each category of Moran's I values
custom_palette_m <- c("#2B3D4F", "#4F628E", "#6C7691", "#A4838D", "#882D43", "#C66B6E")  # Adjust colors if needed

# Map for Moran's I values
map_plot_m <- ggplot(data = iowa_counties2) +
  geom_sf(aes(fill = Ii_Bin), color = "gray30", size = 0.1) +  # Add a thin gray border around counties
  scale_fill_manual(values = setNames(custom_palette_m, labels_m), name = "Moran's I") +  # Use custom color palette
  theme_minimal() +
  theme(legend.position = "left",  # Move legend to the bottom
        legend.key.height = unit(0.5, "cm"),  # Increase legend key height
        legend.text = element_text(size = 6),  # Adjust legend text size
        plot.title = element_text(size = 8, face = "bold", hjust = 0.5)) +  # Adjust title
  labs(title = "Moran's I for Corn Yield in Iowa")  # Adjust title if needed

# Combine the two maps
combined_plot <- grid.arrange(map_plot_m, map_plot_p, ncol = 2)

# Save the combined plot
ggsave("combined_maps.png", plot = combined_plot, width = 12, height = 12, dpi = 300)

```

#Getis Ord's G :

```{r}
library(dplyr)
library(spdep)
library(sf)
library(ggplot2)

# Assuming 'final_data' is your data frame with 'Corn_Yield' and 'County' columns
# and 'iowa_counties' is your sf object with spatial information

# Step 1: Aggregate corn yield by county across all years
aggregated_data <- final_data %>%
  group_by(County) %>%
  summarize(Total_Corn_Yield = sum(Corn_Yield, na.rm = TRUE)) %>%
  arrange(County)  # Ensure that the order of rows matches the spatial weights matrix

# Step 2: Calculate Getis-Ord's G
lw1 <- nb2listw(neighbors, style="W")
getis_ords_g <- localG(aggregated_data$Total_Corn_Yield, lw1)

# Convert getis_ords_g to a data frame and add a column with the County names
getis_ords_g_df <- as.data.frame(getis_ords_g)
getis_ords_g_df$County <- aggregated_data$County

# Step 3: Merge Getis-Ord's G results with the sf object
iowa_counties2 <- iowa_counties %>%
  left_join(getis_ords_g_df, by = c("County"="County"))

# Define breaks and labels for visualization
breaks_m <- c(-Inf, -0.5, -0.25, 0, 0.25, 0.5, Inf)

# Define labels for each bin
labels_m <- c("Very Negative", "Negative", "Slightly Negative", "Slightly Positive", "Positive", "Very Positive")

# Create a new variable representing the bins
iowa_counties2$x<-as.numeric(iowa_counties2$x)
iowa_counties2$G_bin <- cut(iowa_counties2$x, breaks = breaks_m, labels = labels_m)

# Define a custom color palette for each category
custom_palette <- c("#2B3D4F", "#4F628E", "#6C7691", "#A4838D", "#882D43", "#C66B6E")

# Plotting the results
map_plot <- ggplot(data = iowa_counties2) +
  geom_sf(aes(fill = G_bin), color = "gray30", size = 0.1) +
  scale_fill_manual(values = setNames(custom_palette, labels_m), name = "Getis-Ord's G") +
  theme_minimal() +
  theme(legend.position = "left",
        legend.key.height = unit(0.5, "cm"),
        legend.text = element_text(size = 6),
        plot.title = element_text(size = 8, face = "bold", hjust = 0.5)) +
  labs(title = "Getis-Ord's G for Corn Yield in Iowa")

# Save the plot
ggsave("getis_ords_g.png", plot = map_plot, width = 6, height = 5, dpi = 300)

```


```{r}
map_plot
```
```{r}

library(gridExtra)


getis_ords_g <- localG_perm(aggregated_data$Total_Corn_Yield, lw1, nsim = 100, zero.policy = TRUE)

# Convert Getis-Ord's G results to a data frame
getis_ords_g_df <- as.data.frame(attr(getis_ords_g, "internals"))
getis_ords_g_df$County <- aggregated_data$County

# Step 3: Merge Getis-Ord's G results with the sf object
iowa_counties3 <- iowa_counties %>%
  left_join(getis_ords_g_df, by = c("County" = "County"))


# Define breaks for binning the p-values
breaks_p <- c(-Inf, 0.05, Inf)
labels_p <- c("Less than 0.05", "Above 0.05")

# Create a new variable representing the bins for p-values
iowa_counties3$`Pr(z != E(Gi))` <- as.numeric(iowa_counties3$`Pr(z != E(Gi))`)
iowa_counties3$P_Value_Bin <- cut(iowa_counties3$`Pr(z != E(Gi))`, breaks = breaks_p, labels = labels_p)

# Define a custom color palette for each category of p-values
custom_palette_p <- c("#2B3D4F", "#C66B6E")

# Plotting the results for p-values
map_plot_p <- ggplot(data = iowa_counties3) +
  geom_sf(aes(fill = P_Value_Bin), color = "gray30", size = 0.1) +
  scale_fill_manual(values = setNames(custom_palette_p, labels_p), name = "P-Value Getis") +
  theme_minimal() +
  theme(legend.position = "left",
        legend.key.height = unit(0.5, "cm"),
        legend.text = element_text(size = 6),
        plot.title = element_text(size = 8, face = "bold", hjust = 0.5)) +
  labs(title = "P-Value of Getis-Ord's G for Corn Yield in Iowa")

# Combine the two plots side by side
combined_plot <- grid.arrange(map_plot, map_plot_p, ncol = 2)

# Save the combined plot
ggsave("getis_ords_g_combined.png", plot = combined_plot, width = 12, height = 5, dpi = 300)

```

### LISA decomposition:

```{r}
library(dplyr)
library(spdep)
library(sf)
library(ggplot2)

# Step 1: Aggregate corn yield by county across all years
aggregated_data <- final_data %>%
  group_by(County) %>%
  summarize(Total_Corn_Yield = sum(Corn_Yield, na.rm = TRUE))

# Ensure that the order of rows in aggregated_data matches the spatial weights matrix nb
aggregated_data <- aggregated_data %>%
  arrange(County)

# Step 2: Calculate Local Moran's I
lw1 <- nb2listw(neighbors, style="W")
local_morans_i <- localmoran(aggregated_data$Total_Corn_Yield, lw1)

# Convert local_morans_i to a data frame
local_morans_i_df <- as.data.frame(local_morans_i)

# Add a column with the County names
local_morans_i_df$County <- aggregated_data$County

# Extract quadrants
# Assuming that the quadrant information is stored as an attribute of the matrix
quadrants <- attributes(local_morans_i)$quadr

# Convert quadrants to a data frame
quadrants_df <- as.data.frame(quadrants)
quadrants_df$County <- aggregated_data$County

# Step 3: Merge Local Moran's I results with the sf object
iowa_counties2 <- iowa_counties %>%
  left_join(quadrants_df, by = c("County" = "County"))

# Step 4: Define color palette for the quadrants
quadrant_colors <- c("High-High" = "#2B3D4F", "High-Low" = "#4F628E", "Low-High" = "#6C7691", "Low-Low" = "#A4838D")

# Step 5: Map the LISA results
lisa_plot <- ggplot(data = iowa_counties2) +
  geom_sf(aes(fill = factor(pysal)), color = "gray30", size = 0.1) +  # Assuming pysal contains the quadrant information
  scale_fill_manual(values = quadrant_colors, name = "LISA Categories") +  # Use custom color palette
  theme_minimal() +
  theme(legend.position = "left",  # Move legend to the left
        legend.key.height = unit(0.7, "cm"),  # Increase legend key height
        legend.text = element_text(size = 6),  # Adjust legend text size
        plot.title = element_text(size = 8, face = "bold", hjust = 0.5)) +  # Adjust title
  labs(title = "LISA (Local Moran's I) for Corn Yield in Iowa")

print(lisa_plot)

```

combine LISA and p values :

```{r}
combined_plot2<-grid.arrange(lisa_plot,map_plot,  ncol = 1)
```



```{r}
ggsave("combined_plot2.png", plot = combined_plot2, width = 8, height = 5, dpi = 300)

```



#Modeling Phase :
#Preparing the data for modeling:

```{r}
library(INLA)
library(RColorBrewer)
library(tmap)
library(patchwork)
```


```{r}
### preparing the INLA data frame with O, E, 3 ids
S <- length(unique(final_data$County))
T <- length(unique(final_data$YEAR))

t.from <- min(final_data$YEAR)
t.to <- max(final_data$YEAR)


Data <- final_data[order(final_data$YEAR,final_data$County),]

Data.INLA <- data.frame(O = Data$Corn_Yield,
                        E = Data[, c("Temp","precipitation","Hum","dlh","wind","DNI")],
                        ID.area = rep(1:S,T),
                        ID.year = rep(1:T,each=S),
                        ID.area.year = seq(1,T*S))

### Neighboring structure
g <- inla.read.graph("iowa.graph")

Q.xi <- matrix(0, g$n, g$n)
for (i in 1:g$n){
  Q.xi[i,i]=g$nnbs[[i]]
  Q.xi[i,g$nbs[[i]]]=-1
}

Q.Leroux <- diag(S)-Q.xi

### temporal structure matrix 
D1 <- diff(diag(T),differences=1)
Q.gammaRW1 <- t(D1)%*%D1

D2 <- diff(diag(T),differences=2)
Q.gammaRW2 <- t(D2)%*%D2


##  Define appropriate hyperprior distributions using the "expression()" function   
##	- Unif(0,Inf) for standard deviations							                              
##	- Unif(0,1) for the spatial smoothing parameter	

sdunif="expression:
  logdens=-log_precision/2;
  return(logdens)"                

lunif = "expression:
    a = 1;
    b = 1;
    beta = exp(theta)/(1+exp(theta));
    logdens = lgamma(a+b)-lgamma(a)-lgamma(b)+(a-1)*log(beta)+(b-1)*log(1-beta);
    log_jacobian = log(beta*(1-beta));
    return(logdens+log_jacobian)"


### posterior patterns distribution
compute.patterns <- TRUE  ## Set compute.patterns=FALSE if posterior patterns are not required

if(compute.patterns){
  source("posterior_lincombs.R")
}else{
  all.lc <- NULL
}

strategy <- "simplified.laplace"  ## Set strategy="simplified.laplace" for more accurate  results ("gaussian")
```


we are considering the models Type 4 with RW1:
### Type IV interaction and RW1 prior for time 

#Model 3 :

```{r}
R <- kronecker(Q.gammaRW1,Q.xi)
r.def <- S+T-1
A1 <- kronecker(matrix(1,1,T),diag(S))
A2 <- kronecker(diag(T),matrix(1,1,S))
A.constr <- rbind(A1[-1,],A2[-1,])

f.TypeIV.RW1_1 <- O ~ E.precipitation+E.Temp+E.Hum+E.dlh+E.wind+E.DNI+f(ID.area, model="generic1", Cmatrix=Q.Leroux, constr=TRUE,hyper=list(prec=list(prior=sdunif),beta=list(prior=lunif))) +
  f(ID.year, model="rw1", constr=TRUE,
    hyper=list(prec=list(prior=sdunif))) +
  f(ID.area.year, model="generic0", Cmatrix=R, rankdef=r.def,
    constr=TRUE, hyper=list(prec=list(prior=sdunif)),
    extraconstr=list(A=A.constr, e=rep(0,S+T-2)))

TypeIV.RW1_1 <- inla(f.TypeIV.RW1_1, family="gaussian", data=Data.INLA, 
                   control.predictor=list(compute=TRUE, cdf=c(log(1))),
                   control.compute=list(dic=TRUE, cpo=TRUE, waic=TRUE),
                   lincomb=all.lc,control.family=list(hyper=list(prec=list(prior="gaussian", param=c(1,0.01)))),
                   control.inla=list(strategy=strategy))
```

# Model 1 :

```{r}

R <- kronecker(Q.gammaRW1,Q.xi)
r.def <- S+T-1
A1 <- kronecker(matrix(1,1,T),diag(S))
A2 <- kronecker(diag(T),matrix(1,1,S))
A.constr <- rbind(A1[-1,],A2[-1,])

f.TypeIV.RW1_2 <- O ~ E.precipitation+E.Temp+E.Hum+E.dlh+E.wind+E.DNI+f(ID.area, model="generic1", Cmatrix=Q.Leroux, constr=TRUE,hyper=list(prec=list(prior=sdunif),beta=list(prior=lunif))) +
  f(ID.year, model="rw1", constr=TRUE,
    hyper=list(prec=list(prior=sdunif))) +
  f(ID.area.year, model="generic0", Cmatrix=R, rankdef=r.def,
    constr=TRUE, hyper=list(prec=list(prior=sdunif)),
    extraconstr=list(A=A.constr, e=rep(0,S+T-2)))

TypeIV.RW1_2 <- inla(f.TypeIV.RW1_2, family="gaussian", data=Data.INLA, 
                   control.predictor=list(compute=TRUE, cdf=c(log(1))),
                   control.compute=list(dic=TRUE, cpo=TRUE, waic=TRUE),
                   lincomb=all.lc
                   ,control.family=list(hyper=list(prec=list(prior="loggamma", param=c(1,0.01)))),
                   control.inla=list(strategy=strategy))
```

#Model 2 :
```{r}

R <- kronecker(Q.gammaRW1,Q.xi)
r.def <- S+T-1
A1 <- kronecker(matrix(1,1,T),diag(S))
A2 <- kronecker(diag(T),matrix(1,1,S))
A.constr <- rbind(A1[-1,],A2[-1,])

f.TypeIV.RW1_3 <- O ~ E.precipitation+E.Temp+E.Hum+E.dlh+E.wind+E.DNI+f(ID.area, model="generic1", Cmatrix=Q.Leroux, constr=TRUE,hyper=list(prec=list(prior=sdunif),beta=list(prior=lunif))) +
  f(ID.year, model="rw1", constr=TRUE,
    hyper=list(prec=list(prior=sdunif))) +
  f(ID.area.year, model="generic0", Cmatrix=R, rankdef=r.def,
    constr=TRUE, hyper=list(prec=list(prior=sdunif)),
    extraconstr=list(A=A.constr, e=rep(0,S+T-2)))

TypeIV.RW1_3 <- inla(f.TypeIV.RW1_3, family="gaussian", data=Data.INLA, 
                   control.predictor=list(compute=TRUE, cdf=c(log(1))),
                   control.compute=list(dic=TRUE, cpo=TRUE, waic=TRUE),
                   lincomb=all.lc,
                   control.family=list(hyper=list(prec=list(prior="loggamma",param = c(0.01, 0.01)))),
                   control.inla=list(strategy=strategy))


```


#Model 4:

```{r}
R <- kronecker(Q.gammaRW1,Q.xi)
r.def <- S+T-1
A1 <- kronecker(matrix(1,1,T),diag(S))
A2 <- kronecker(diag(T),matrix(1,1,S))
A.constr <- rbind(A1[-1,],A2[-1,])

f.TypeIV.RW1_7 <- O ~ E.precipitation+E.Temp+E.Hum+E.dlh+E.wind+E.DNI+f(ID.area, model="generic1", Cmatrix=Q.Leroux, constr=TRUE,hyper=list(prec=list(prior=sdunif),beta=list(prior=lunif))) +
  f(ID.year, model="rw1", constr=TRUE,
    hyper=list(prec=list(prior=sdunif))) +
  f(ID.area.year, model="generic0", Cmatrix=R, rankdef=r.def,
    constr=TRUE, hyper=list(prec=list(prior=sdunif)),
    extraconstr=list(A=A.constr, e=rep(0,S+T-2)))

TypeIV.RW1_7 <- inla(f.TypeIV.RW1_7, family="gaussian", data=Data.INLA, 
                   control.predictor=list(compute=TRUE, cdf=c(log(1))),
                   control.compute=list(dic=TRUE, cpo=TRUE, waic=TRUE),
                   lincomb=all.lc,
                   control.family=list(hyper=list(prec=list(prior="gaussian",param = c(0.01, 0.01)))),
                   control.inla=list(strategy=strategy))


```


Choosing the best model among them :

```{r}
MODELS <- list(Model3=TypeIV.RW1_1,
               Model1=TypeIV.RW1_2,
               Model2=TypeIV.RW1_3,
               Model4=TypeIV.RW1_7
               )

DIC <- function(x){
  data.frame(mean.deviance=x$dic$mean.deviance, ## posterior mean deviance
             p.eff=x$dic$p.eff,                 ## effective number of parameters
             DIC=x$dic$dic,                     ## Deviance Information Criterion
             WAIC=x$waic$waic,                  ## Watanabe-Akaike information criterion
             LS=-sum(log(x$cpo$cpo)),           ## Logarithmic Score (see inla.cpo function)
             Time=x$cpu.used[4])
}

do.call(rbind,lapply(MODELS, DIC))
```

###Selected model parameters and analysis: 

```{r}
Model <- TypeIV.RW1_1


```



```{r}
print(summary(Model))

```

```{r}
# Extracting mean and standard deviation for fixed effect
fixed_summary <- summary(Model)$fixed
alpha_mean <- fixed_summary[1, "mean"]
alpha_sd <- fixed_summary[1, "sd"]

# Calculating hyperparameter statistics
sigma2_S_mean <- inla.emarginal(function(x) 1/x, Model$marginals.hyperpar$"Precision for ID.area")
sigma2_S_sd <- sqrt(inla.emarginal(function(x) 1/(x^2), Model$marginals.hyperpar$"Precision for ID.area") - sigma2_S_mean^2)

lambda_S_mean <- Model$summary.hyperpar[2, 1]
lambda_S_sd <- Model$summary.hyperpar[2, 2]

sigma2_T_mean <- inla.emarginal(function(x) 1/x, Model$marginals.hyperpar$"Precision for ID.year")
sigma2_T_sd <- sqrt(inla.emarginal(function(x) 1/(x^2), Model$marginals.hyperpar$"Precision for ID.year") - sigma2_T_mean^2)

sigma2_ST_mean <- inla.emarginal(function(x) 1/x, Model$marginals.hyperpar$"Precision for ID.area.year")
sigma2_ST_sd <- sqrt(inla.emarginal(function(x) 1/(x^2), Model$marginals.hyperpar$"Precision for ID.area.year") - sigma2_ST_mean^2)

# Combine results into a data frame
table <- data.frame(
  par = c("alpha", "sigma2_S", "lambda_S", "sigma2_T", "sigma2_ST"),
  mean.model = c(alpha_mean, sigma2_S_mean, lambda_S_mean, sigma2_T_mean, sigma2_ST_mean),
  sd.model = c(alpha_sd, sigma2_S_sd, lambda_S_sd, sigma2_T_sd, sigma2_ST_sd)
)

print(table)
```


```{r}
summary(TypeIV.RW1_1$summary.fitted.values)
```



### Percentage of explained variability by the spatial, temporal and spatio-temporal patterns

```{r}
alpha <- Model$summary.lincomb.derived$mean[1]
risks.S <- matrix(Model$summary.lincomb.derived$mean[2:(S+1)],S,T)
risks.T <- t(matrix(Model$summary.lincomb.derived$mean[(S+2):(S+T+1)],T,S))
risks.ST <- matrix(Model$summary.lincomb.derived$'mean'[(S+T+2):(S+T+S*T+1)], S, T)

varS <-var(as.vector(risks.S))
varT <- var(as.vector(risks.T))
varST <- var(as.vector(risks.ST))
c(varS,varT,varST)/(varS+varT+varST)
```

